---
title: "Using GitHub Copilot Coding Agents"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using GitHub Copilot Coding Agents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(demos)
library(gtsummary)
```

# Introduction

GitHub Copilot is an AI-powered coding assistant that can help you write, understand, and review code. Beyond line-by-line autocomplete, Copilot also offers **coding agents** — autonomous AI assistants that can plan and execute multi-step coding tasks on your behalf, directly within GitHub issues and pull requests.

This vignette walks through how to use Copilot's coding agents and demonstrates the kind of analysis they can help you produce.

---

# What Are Copilot Coding Agents?

Copilot coding agents (also called **Copilot Workspace** or **agent mode** in GitHub) allow you to describe a task in natural language — such as in a GitHub Issue — and have Copilot automatically:

1. Plan the steps needed to complete the task.
2. Write or modify code files.
3. Open a pull request with the proposed changes.

You remain in control: you can review, edit, and approve every change before it is merged.

---

# Step-by-Step: Using Copilot Coding Agents

## Step 1 — Open a GitHub Issue

Describe the task you want Copilot to perform in plain English. Be specific about inputs, outputs, and any constraints. For example:

> *Create a vignette with step-by-step instructions on using Copilot's coding agents. Include a sample analysis using the mtcars data. Summarize mpg, disp, wt, and cyl by am. Fit a linear model (mpg = disp + wt + cyl), and display the model estimates in a table.*

## Step 2 — Assign the Issue to Copilot

In the GitHub issue sidebar, click **Assignees** and select **Copilot** from the list. This triggers the agent to begin working on the issue.

## Step 3 — Review the Agent's Plan

Copilot will post a comment outlining its plan — the files it intends to create or modify and the steps it will take. Review this plan and add any clarifying comments if needed.

## Step 4 — Copilot Opens a Pull Request

Once the agent has implemented the changes, it will open a pull request linked to the issue. The PR will contain all new or modified files.

## Step 5 — Review, Iterate, and Merge

Review the pull request diff. You can:

- Leave review comments asking for revisions (Copilot will update the PR).
- Edit files directly in the PR branch.
- Approve and merge when satisfied.

---

# Sample Analysis: mtcars

The following analysis was produced as a demonstration of the work Copilot coding agents can generate.

## Summary Statistics by Transmission Type

The `am` variable in `mtcars` encodes transmission type: `0` = automatic, `1` = manual. We summarize `mpg`, `disp`, `wt`, and `cyl` by `am` using the {gtsummary} package.

```{r summary-table, message=FALSE, warning=FALSE}
mtcars |>
  gtsummary::tbl_summary(
    by = am,
    include = c(mpg, disp, wt, cyl),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 2,
    label = list(
      mpg  ~ "Miles per Gallon",
      disp ~ "Displacement (cu.in.)",
      wt   ~ "Weight (1000 lbs)",
      cyl  ~ "Number of Cylinders"
    )
  ) |>
  gtsummary::modify_header(
    stat_1 = "**Automatic**\nN = {n}",
    stat_2 = "**Manual**\nN = {n}"
  ) |>
  gtsummary::modify_spanning_header(
    gtsummary::all_stat_cols() ~ "**Transmission**"
  ) |>
  gtsummary::add_overall() |>
  gtsummary::add_p() |>
  gtsummary::bold_labels()
```

Vehicles with manual transmissions (`am = 1`) tend to have higher fuel efficiency, lower displacement, and lower weight compared to automatic transmission vehicles.

---

## Linear Model: mpg ~ disp + wt + cyl

We fit a linear regression model predicting `mpg` from displacement (`disp`), weight (`wt`), and number of cylinders (`cyl`).

```{r linear-model, message=FALSE, warning=FALSE}
fit <- lm(mpg ~ disp + wt + cyl, data = mtcars)

gtsummary::tbl_regression(
  x = fit,
  label = list(
    disp ~ "Displacement (cu.in.)",
    wt   ~ "Weight (1000 lbs)",
    cyl  ~ "Number of Cylinders"
  )
) |>
  gtsummary::add_glance_source_note(
    include = c(r.squared, adj.r.squared, nobs)
  ) |>
  gtsummary::bold_labels() |>
  gtsummary::bold_p(t = 0.05) |>
  gtsummary::modify_caption("**Linear Model: mpg ~ disp + wt + cyl**")
```

The model estimates show how each predictor is associated with fuel efficiency after adjusting for the other variables. Weight (`wt`) and number of cylinders (`cyl`) are typically the strongest predictors of `mpg` in this dataset.

---

# Tips for Getting the Best Results from Copilot Coding Agents

- **Be specific in your issue description.** The more detail you provide about expected inputs, outputs, and coding style, the better the agent's output will be.
- **Iterate through PR review comments.** If the generated code isn't quite right, leave a targeted comment and the agent will revise.
- **Use Copilot Chat for smaller tasks.** For quick, in-editor questions or snippets, Copilot Chat (available in RStudio via the GitHub Copilot extension or VS Code) is faster than the full agent workflow.
- **Always review generated code.** Copilot agents are powerful, but human review ensures correctness, security, and adherence to project conventions.
