---
title: "Demo of the gtsummary package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Demo of the gtsummary package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(demos)
library(gtsummary)
```

# Introduction

The {gtsummary} package provides an elegant and flexible way to create publication-ready analytical and summary tables using the R programming language. The {gtsummary} package summarizes data sets, regression models, and more, using sensible defaults with highly customizable capabilities ([ref.](https://www.danieldsjoberg.com/gtsummary/)).  

I've found this package to be more customizable and flexible than {arsenal} or {table1}. It also renders nicely to Word documents through the utilization of the {flextable} package.

# [Summarize data frames or tibbles](https://www.danieldsjoberg.com/gtsummary/articles/tbl_summary.html)

## Basics  

The `tbl_summary()` function is the main function to generate a summary table of your data frame. A more detailed tutorial can be found [here](https://www.danieldsjoberg.com/gtsummary/articles/tbl_summary.html).

<br>

```{r echo=TRUE}
tbl_summary(
  data = mtcars,
  by = NULL,
  label = NULL,
  statistic = list(
    all_continuous() ~ "{median} ({p25}, {p75})", 
    all_categorical() ~ "{n} ({p}%)"
  ),
  digits = NULL,
  type = NULL,
  value = NULL,
  missing = c("ifany", "no", "always"),
  missing_text = "Unknown",
  missing_stat = "{N_miss}",
  sort = all_categorical(FALSE) ~ "alphanumeric",
  percent = c("column", "row", "cell"), # default = column
  include = everything()
)
```

<br>

{gtsummary} will report median (Q1, Q3) for continuous variables and n (%) for character/factor variables. If a numeric variable only has a few discrete values, it will automatically treat that variable as a factor.  

The summary statistics can be modified through the `type` argument (e.g. continuous or categorical) and the `statistic` argument. The `statistic` argument value are interpreted using `glue::glue()` syntax (e.g. the R expressions wrapped in curly braces "{mean} ({sd})").

To generate summaries by group, set the `by` argument to your variable of interest. The header labels can be changed using the `modify_header()` function, and a spanning header to label the by variable can be added using the `modify_spanning_header()` function.

<br>

```{r}
gtsummary::tbl_summary(data = mtcars, by = am) |>
  gtsummary::modify_header(
    stat_1 = "**Automatic**</br>N = {n} ({style_percent(p)}%)",
    stat_2 = "**Manual**</br>N = {n} ({style_percent(p)}%)") |>
  gtsummary::modify_spanning_header(gtsummary::all_stat_cols() ~ "Transmission")
```

<br>

## Add Columns

By design, the `tbl_sumamry()` function only displays the summary statistics for your variables. Through the use of `add_*` functions, you can add certain columns to the table such as overall summary statistic, p-values, q-values, or confidence intervals.

<ul>
  <li>`add_overall()`: Adds a column with overall summary statistics to tables.</li>
  <li>`add_p()`: Adds p-values to tables created by `tbl_summary()` by comparing values across groups.
    <ul>
      <li>Extend with `separate_p_footnotes()` for each statistical method to have its own footnote.</li>
      <li>A list of available tests/methods and an example of creating custom functions can be found [here](https://www.danieldsjoberg.com/gtsummary/reference/tests.html#tbl-summary-gt-add-p-).</li>
    </ul>
  </li>
  <li>`add_q()`: Adjustments to p-values are performed with `stats::p.adjust()`.</li>
  <li>`add_ci()`: Add a new column with the confidence intervals for proportions, means, etc.</li>
</ul>

<br>

```{r message=FALSE, warning=FALSE}
gtsummary::tbl_summary(data = mtcars, by = am) |>
  gtsummary::add_overall() |>
  gtsummary::add_p() |>
  gtsummary::separate_p_footnotes() |>
  gtsummary::add_q() |>
  gtsummary::modify_header(
    stat_1 = "**Automatic**</br>N = {n} ({style_percent(p)}%)",
    stat_2 = "**Manual**</br>N = {n} ({style_percent(p)}%)") |>
  gtsummary::modify_spanning_header(gtsummary::all_stat_cols() ~ "Transmission")
  
```

<br>

## Strata

If you need to generate summaries stratified by a third variable, this can be done using the `tbl_strata()` function.

<br>

```{r}
tbl_strata(
  data = mtcars,
  strata = vs,
  .tbl_fun = ~ .x |>
    gtsummary::tbl_summary(by = am)#,
  # ...,
  # .sep = ", ",
  # .combine_with = c("tbl_merge", "tbl_stack"),
  # .combine_args = NULL,
  # .header = ifelse(.combine_with == "tbl_merge", "**{strata}**", "{strata}"),
  # .stack_group_header = NULL,
  # .quiet = NULL
)
```

# [Summarize Regression Models](https://www.danieldsjoberg.com/gtsummary/articles/tbl_regression.html)

To summarize the results of a regression model, you can pass the fitted model to the `tbl_regression()` function. The function uses `broom::tidy` behind the scenes to generate the initial model formatting. The list of supported models can be found [here](https://www.danieldsjoberg.com/gtsummary/articles/tbl_regression.html#supported-models), note that some model types require additional package dependencies (e.g. {broom.mixed}).

<br>

```{r}
fit_lm <- lm(mpg ~ hp + as.factor(am) + as.factor(cyl), data = mtcars)

gtsummary::tbl_regression(
  x = fit_lm #,
  # label = NULL,
  # exponentiate = FALSE,
  # include = everything(),
  # show_single_row = NULL,
  # conf.level = 0.95,
  # intercept = FALSE,
  # estimate_fun = ifelse(exponentiate, label_style_ratio(), label_style_sigfig()),
  # pvalue_fun = label_style_pvalue(digits = 1),
  # tidy_fun = broom.helpers::tidy_with_broom_or_parameters,
  # add_estimate_to_reference_rows = FALSE,
  # conf.int = TRUE,
  # ...
)
```

<br>

```{r}
fit_glm <- glm(as.factor(am) ~ mpg + as.factor(vs), data = mtcars, family = binomial)
gtsummary::tbl_regression(fit_glm)
gtsummary::tbl_regression(fit_glm, exponentiate = TRUE)
```

<br>

## Helpers

The {gtsummary} package has built-in functions for adding to results from `tbl_regression()`. The following functions add columns and/or information to the regression table.

<ul>
  <li>`add_global_p()`: adds the global p-value for categorical variables (uses `car::Anova()`)</li>
  <li>`add_glance_source_note()`: adds statistics from `broom::glance()` as source note
    <ul>
      <li>`broom::glance()` returns statistics such as $R^2$, $\text{Adjusted }R^2$, degrees of freedom, AIC, BIC, deviance, and residual degrees of freedom.</li>
      <li>The statistics returned by `broom::glance()` will vary based on the model.</li>
    </ul>
  </li>
  <li>`add_vif()`: adds column of the variance inflation factors (VIF)</li>
  <li>`add_q()`: add a column of q values to control for multiple comparisons.</li>
</ul>

<br>

The {gtsummary} package also has additional helpers to modify the appearance of the table.

<ul>
  <li>`modify_header()`: updated column headers.</li>
  <li>`modify_footnote()`: updated column footnote.</li>
  <li>`modify_spanning_header()`: update the spanning headers.</li>
  <li>`modify_caption()`: update table caption/title.</li>
  <li>`bold_labels()`: bold variable labels.</li>
  <li>`bold_levels()`: bold variable levels.</li>
  <li>`italicize_labes()/italicize_levels()`: italicize variable lables/levels.</li>
  <li>`bold_p()`: bold <b>significant</b> p-values.</li>
</ul>


```{r}
gtsummary::tbl_regression(fit_lm) |>
  gtsummary::add_global_p() |>
  gtsummary::bold_labels() |>
  gtsummary::italicize_levels() |>
  gtsummary::bold_p()
```

# [Report results inline](https://www.danieldsjoberg.com/gtsummary/articles/inline_text.html#introduction)

Reproducible reports are an important part of good practices. We often need to report the results from a table in the text of an R markdown report. Inline reporting has been made simple with `inline_text()`. The `inline_text()` function reports statistics from {gtsummary} tables inline in an R markdown report.

```{r}
tab1 <- tbl_summary(trial, by = trt, include = c(marker, stage))
tab1
```

```{r, echo=FALSE, eval=TRUE}
cat('The median (IQR) marker level in the Drug A and Drug B groups are `r inline_text(tab1, variable = marker, column = "Drug A")` and `r inline_text(tab1, variable = marker, column = "Drug B")`, respectively.')
```

The median (IQR) marker level in the Drug A and Drug B groups are `r inline_text(tab1, variable = marker, column = "Drug A")` and `r inline_text(tab1, variable = marker, column = "Drug B")`, respectively.

