---
title: "Joinpoint Analysis: Modeling and Analyzing Trends in Rates Over Time"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
    theme: default
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(ggplot2)
library(readr)
```

## Outline

- What is Joinpoint?\
- Example: Cancer Incidence Rates\
- Fitting a Model in R\
- Exporting Results for Joinpoint\
- Joinpoint Software Overview\
- Interpreting Results (APC & AAPC)

## What is Joinpoint?

**Joinpoint Trend Analysis Software** is a free program developed by the
National Cancer Institute (NCI).

- Fits **joinpoint regression** — piecewise linear models on a log scale
- Identifies points in time where the trend changes significantly (**joinpoints**)
- Estimates **Annual Percent Change (APC)** for each trend segment
- Estimates **Average Annual Percent Change (AAPC)** across the entire period

::: aside
Download: <https://surveillance.cancer.gov/joinpoint/>
:::

## Joinpoint Regression

For a rate $r(t)$ observed over years $t$, the model is:

$$\log[r(t)] = \beta_0 + \beta_1 t + \delta_1(t - \tau_1)_+ + \cdots + \delta_k(t - \tau_k)_+$$

where $\tau_1, \ldots, \tau_k$ are the **joinpoints** (change points) and
$(x)_+ = \max(0,\, x)$.

The APC for a segment is:

$$\text{APC} = (e^{\beta} - 1) \times 100\%$$

## When Should You Use Joinpoint?

Use Joinpoint when you want to:

- Describe **temporal trends** in incidence, mortality, or prevalence rates
- Identify **when** a trend changed direction or pace
- Summarize a long time series with a single **AAPC** statistic
- Produce results consistent with **NCI SEER** publications

## Example Data {.smaller}

Simulated annual cancer incidence counts and population sizes (2000–2022):

```{r}
#| label: create-data
#| echo: true

set.seed(42)
years <- 2000:2022

# Simulate population (growing ~1% per year)
population <- round(500000 * 1.01^(years - 2000))

# Simulate true rate per 100,000 with a joinpoint around 2010
true_rate <- ifelse(
  years <= 2010,
  50 * exp(-0.02 * (years - 2000)),   # declining trend 2000–2010
  50 * exp(-0.02 * 10) * exp(0.01 * (years - 2010))  # rising trend 2010–2022
)

# Simulate observed counts (Poisson)
counts <- rpois(length(years), lambda = true_rate * population / 1e5)

cancer_data <- tibble(
  year       = years,
  population = population,
  count      = counts,
  rate       = count / population * 1e5
)

head(cancer_data)
```

## Visualizing the Raw Rates

```{r}
#| label: fig-raw-rates
#| fig-cap: "Simulated cancer incidence rates per 100,000 (2000–2022)"
#| echo: false

ggplot(cancer_data, aes(x = year, y = rate)) +
  geom_point(size = 2, color = "#2C3E82") +
  geom_line(color = "#2C3E82", linewidth = 0.8) +
  labs(
    x = "Year",
    y = "Rate per 100,000",
    title = "Cancer Incidence Rate Over Time"
  ) +
  theme_bw(base_size = 14)
```

## Fitting a Poisson Regression Model in R {.smaller}

A **Poisson regression** with `log(population)` as an offset is the standard
approach for modeling count data and extracting rates:

```{r}
#| label: fit-model
#| echo: true

# Fit Poisson regression: count ~ year, with log(population) offset
poisson_model <- glm(
  count ~ year,
  data   = cancer_data,
  family = poisson(link = "log"),
  offset = log(population)
)

summary(poisson_model)
```

## Extracting Fitted Rates and Standard Errors {.smaller}

Joinpoint needs the **rate** and its **standard error** for each year.

```{r}
#| label: extract-se
#| echo: true

# Predicted log-rate and its standard error from the model
pred <- predict(poisson_model, type = "link", se.fit = TRUE)

cancer_data <- cancer_data |>
  mutate(
    fitted_log_rate = pred$fit,
    se_log_rate     = pred$se.fit,
    fitted_rate     = exp(fitted_log_rate) * 1e5,
    # SE of the rate on the original scale (delta method)
    se_rate         = fitted_rate * se_log_rate
  )

cancer_data |>
  select(year, count, population, rate, fitted_rate, se_rate) |>
  head(8)
```

## Why Use Fitted Values? {.smaller}

::: columns
::: {.column width="48%"}
**Raw rates** can be noisy, especially for:

- Small populations
- Rare outcomes
- Short time windows

:::
::: {.column width="48%"}
**Model-smoothed rates** provide:

- Stable estimates of the underlying trend
- A principled standard error for each time point
- Consistency with methods used in cancer surveillance
:::
:::

> Joinpoint can accept either raw rates + standard errors **or**
> raw counts + population. The smoothed approach is useful when
> additional covariates (age, sex, race) are adjusted in R.

## Preparing the Joinpoint Export File {.smaller}

Joinpoint expects a **tab-delimited text file** with at minimum:

| Column | Description |
|--------|-------------|
| `Year` | Calendar year (time variable) |
| `Rate` | Rate per 100,000 |
| `SE`   | Standard error of the rate |

Optional columns: `Count`, `Population`, grouping variables.

## Creating the Export File in R {.smaller}

```{r}
#| label: export-joinpoint
#| echo: true

joinpoint_export <- cancer_data |>
  select(
    Year       = year,
    Rate       = fitted_rate,  # model-smoothed rate per 100,000
    SE         = se_rate,      # SE from fitted model (delta method)
    Count      = count,
    Population = population
  )

# Write tab-delimited file (Joinpoint default format)
write_tsv(joinpoint_export, "joinpoint_input.txt")

head(joinpoint_export, 5)
```

::: aside
`write_tsv()` is from the **{readr}** package. The `.txt` extension is
recommended by Joinpoint.
:::

## Joinpoint Input File Format

```
Year    Rate        SE          Count   Population
2000    49.73455    0.2228      249     500000
2001    48.34212    0.2192      244     505000
2002    47.04291    0.2154      239     510050
...
```

Rules:

- **Tab-separated** (not comma-separated)
- **No missing values** in Year or Rate columns
- Column headers must match what you specify in the software
- One row per time point (or per time point × group combination)

## Joinpoint Software: Getting Started

1. Download and install from <https://surveillance.cancer.gov/joinpoint/>
2. Open Joinpoint → **File → New Session**
3. Import your tab-delimited text file
4. Specify:
   - **Time variable**: `Year`
   - **Rate variable**: `Rate`
   - **Standard error variable**: `SE` *(or use count/population)*
5. Set the **maximum number of joinpoints** (typically 0–4)

## Joinpoint Software: Key Settings

::: columns
::: {.column width="55%"}
**Model Selection**

- Default method: **Weighted BIC (WBIC)**
- Fits models with 0, 1, 2, … *k* joinpoints and selects the one with the lowest WBIC
- WBIC penalizes model complexity, balancing fit against the number of joinpoints
- Method: Weighted least squares on log scale
:::
::: {.column width="40%"}
**Additional Options**

- Autocorrelation correction
- Parallelism test across groups
- Confidence intervals via Monte Carlo
:::
:::

## Running the Analysis

1. Click **Run** (or **Ctrl+R**)
2. Joinpoint fits models with 0, 1, 2, … *k* joinpoints
3. **Weighted BIC** selects the best model (lowest WBIC wins)
4. Results appear in the **Output** window

The software reports:

- Segment-specific **APCs** with 95% confidence intervals
- **AAPC** for a specified time period
- Joinpoint locations (years) with 95% CIs

## Interpreting APC {.smaller}

The **Annual Percent Change (APC)** describes the slope within each trend segment:

| APC | Interpretation |
|-----|----------------|
| Positive (+) | Rate is **increasing** |
| Negative (−) | Rate is **decreasing** |
| Not significant | Trend is **stable** |

Example output:

```
Segment 1: 2000-2010  APC = -2.0% (95% CI: -2.4%, -1.6%)  p < 0.001
Segment 2: 2010-2022  APC = +1.0% (95% CI:  0.7%,  1.3%)  p < 0.001
```

This tells us: rates declined ~2%/year through 2010, then rose ~1%/year.

## Interpreting AAPC {.smaller}

The **Average Annual Percent Change (AAPC)** summarizes the entire period as a
weighted average of the segment APCs:

$$\text{AAPC} = \left[\left(\prod_{j=1}^{k+1} (1 + \text{APC}_j/100)^{w_j}\right)^{1/\sum w_j} - 1\right] \times 100\%$$

where $w_j$ is the length of segment $j$.

> **Use AAPC** when you want a single number to describe the overall trend
> (e.g., in an abstract or executive summary).
>
> **Use segment APCs** when the timing of change is scientifically important.

## Exporting Results from Joinpoint

Joinpoint produces several exportable outputs:

- **Results table** (APC, AAPC, joinpoint locations) — export as `.csv` or copy to Excel
- **Graphs** — save as `.emf`, `.png`, or `.pdf`
- **Session file** (`.jps`) — saves all settings for reproducibility

Re-import the results table into R for further analysis or reporting with
packages like **{gtsummary}** or **{flextable}**.

## End-to-End Workflow Summary

```
R                               Joinpoint Software
───────────────────────         ──────────────────────────────────
1. Load / clean data            4. Import tab-delimited file
2. Fit Poisson model            5. Specify variables & settings
3. Export rate + SE  ──────►    6. Weighted BIC model selection
   (write_tsv)                  7. Review APC / AAPC results
                      ◄──────   8. Export results table / graphs
9. Import results               
10. Report / visualize           
```

## Key Takeaways

- **R** handles data preparation, covariate adjustment, and rate estimation
- **Joinpoint** identifies when trends changed and quantifies the rate of change
- The **APC** describes the trend in each segment; the **AAPC** summarizes the whole period
- Provide Joinpoint with the **rate per 100,000** and its **standard error**
- Save your Joinpoint **session file** (`.jps`) for reproducibility

## Resources

- Joinpoint software & documentation:\
  <https://surveillance.cancer.gov/joinpoint/>

- NCI SEER Stat tutorials:\
  <https://seer.cancer.gov/seerstat/>

- Clegg LX et al. (2009). Joinpoint regression analysis of cancer trends.
  *Journal of the National Cancer Institute*, 101(12), 913–914.

- Tiwari RC et al. (2005). Weighted AIC and BIC for joinpoint regression.
  *Statistics and Its Interface*, 2(4), 379–396.
