---
title: "Introduction to renv"
subtitle: "Reproducible R Environments"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
    theme: default
---

## What is renv?

**renv** is an R package that helps you create **reproducible environments** for your R projects.

::: {.incremental}
- Manages R package dependencies on a per-project basis
- Records exact package versions used in your project
- Makes it easy to restore the same environment later
- Similar to Python's virtualenv or Node's npm
:::

::: {.notes}
renv provides isolated, reproducible environments for your R projects
:::

## Why Use renv? {.smaller}

**Problems it solves:**

::: {.incremental}
- **Version conflicts**: Different projects need different package versions
- **Reproducibility**: "It works on my machine" syndrome
- **Collaboration**: Team members need identical environments
- **Time travel**: Restore project state from months/years ago
:::

. . .

**Benefits:**

::: {.incremental}
- Project-specific libraries (isolation)
- Explicit dependency tracking
- Faster package installation (shared cache)
- Works with R packages, Bioconductor, and GitHub packages
:::

## Setting Up renv for a New Project

**Initialize renv in your project:**

```{r}
#| eval: false
#| echo: true

# Install renv (once)
install.packages("renv")

# Initialize renv in your project directory
renv::init()
```

. . .

**What happens:**

::: {.incremental}
- Creates `renv/` folder (project library)
- Creates `renv.lock` file (dependency record)
- Creates `.Rprofile` (activates renv on startup)
- Copies currently installed packages to project library
:::

## Using renv: Basic Workflow

**1. Install packages as usual:**
```{r}
#| eval: false
#| echo: true
install.packages("dplyr")
install.packages("ggplot2")
```

. . .

**2. Take a snapshot:**
```{r}
#| eval: false
#| echo: true
renv::snapshot()
```
Records current package versions to `renv.lock`

. . .

**3. Restore environment (on different machine/later):**
```{r}
#| eval: false
#| echo: true
renv::restore()
```
Installs exact versions from `renv.lock`

## Working with renv: Key Functions {.smaller}

| Function | Purpose |
|----------|---------|
| `renv::init()` | Initialize renv in a project |
| `renv::snapshot()` | Save current package versions to lockfile |
| `renv::restore()` | Install packages from lockfile |
| `renv::status()` | Check synchronization status |
| `renv::update()` | Update packages to latest versions |
| `renv::install()` | Install packages (CRAN, GitHub, Bioconductor) |
| `renv::remove()` | Remove packages from project library |

. . .

**Pro tip:** Run `renv::status()` regularly to see if your lockfile is synchronized with installed packages!

## Collaboration Workflow with renv

**Developer A (setting up project):**
```{r}
#| eval: false
#| echo: true
renv::init()
# ... work on project, install packages ...
renv::snapshot()
# Commit renv.lock to version control (Git)
```

. . .

**Developer B (joining project):**
```{r}
#| eval: false
#| echo: true
# Clone/pull repository
# Open R project - renv activates automatically
renv::restore()  # Install all dependencies
# Start working!
```

. . .

**Key files to commit:**
- ✅ `renv.lock` (dependency record)
- ✅ `.Rprofile` (activation script)
- ✅ `renv/settings.json` (configuration)
- ❌ `renv/library/` (don't commit installed packages)

## Using renv with R Packages

**For R package development:**

```{r}
#| eval: false
#| echo: true
# Initialize renv
renv::init()

# Install development dependencies
renv::install("devtools")
renv::install("testthat")
renv::install("roxygen2")

# Snapshot your development environment
renv::snapshot(type = "all")
```

. . .

**Benefits for package developers:**

::: {.incremental}
- Separate development dependencies from package dependencies
- Consistent testing environment across contributors
- Easier CI/CD setup
:::

## Best Practices and Tips {.smaller}

**Do's:**

::: {.incremental}
- ✅ Commit `renv.lock` to version control
- ✅ Run `renv::snapshot()` after installing/updating packages
- ✅ Use `renv::status()` to check synchronization
- ✅ Run `renv::restore()` after pulling changes
:::

. . .

**Don'ts:**

::: {.incremental}
- ❌ Don't commit the `renv/library/` folder (too large)
- ❌ Don't manually edit `renv.lock` (use renv functions)
- ❌ Don't mix global and project libraries
:::

. . .

**Troubleshooting:**

```{r}
#| eval: false
#| echo: true
# If something goes wrong
renv::diagnostics()  # Check for issues
renv::rebuild()      # Rebuild packages
renv::repair()       # Repair the library
```

## Common Scenarios {.smaller}

**Updating a single package:**
```{r}
#| eval: false
#| echo: true
renv::update("dplyr")
renv::snapshot()
```

. . .

**Installing from GitHub:**
```{r}
#| eval: false
#| echo: true
renv::install("tidyverse/dplyr")
renv::snapshot()
```

. . .

**Rolling back to previous state:**
```{r}
#| eval: false
#| echo: true
renv::history()     # View snapshot history
renv::revert()      # Revert to previous snapshot
```

. . .

**Deactivating renv (if needed):**
```{r}
#| eval: false
#| echo: true
renv::deactivate()  # Use global library instead
```

## Resources and Getting Help

**Documentation:**

- [renv website](https://rstudio.github.io/renv/)
- [Introduction vignette](https://rstudio.github.io/renv/articles/renv.html)
- [Collaboration guide](https://rstudio.github.io/renv/articles/collaborating.html)

. . .

**Quick reference:**

```{r}
#| eval: false
#| echo: true
?renv                  # Package help
vignette("renv")       # Introduction vignette
renv::diagnostics()    # Troubleshooting
```

. . .

**Questions?**

*Start using renv in your next R project for better reproducibility!*
