---
title: "Guide on Code Styling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guide on Code Styling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# [Tidyverse Style Guide](https://style.tidyverse.org/)

## Files

### Internal structure

Use commented lines of `-` and `=` to break up your file into easily readable chunks.

```{r}
# Load data -----------------

# Plot data -----------------

```

Load ALL packages at once at the beginning of the file. This is more transparent than loading packages throughout the code or having hidden dependencies that are loaded in a startup file (`.Rprofile`).

## Syntax

### Spacing

-   Commas  
    -   Always put a space after a comma, never before.  

```{r, eval=FALSE}
# Good
x[, 1]

# Bad
x[,1]
x[ ,1]
x[ , 1]
```

-   Parentheses  
    -   Do not put spaces inside or outside parentheses for regular function calls.
    
```{r, eval=FALSE}
# Good
mean(x, na.rm = TRUE)

# Bad
mean (x, na.rm = TRUE)
mean( x, na.rm = TRUE )

```

    -   Place a space after `()` used for function arguments:
    
```{r, eval=FALSE}
# Good
function(x) {}

# Bad
function (x) {}
function(x){}
```

-   Embracing
    -   The embracing operator `{ }`, should always have inner spaces to help emphasise its special behavior.
    
```{r, eval=FALSE}
# Good
max_by <- function(data, var, by) {
  data |>
    group_by({{ by }}) |>
    summarise(maximum = max({{ var }}, na.rm = TRUE))
}

# Bad
max_by <- function(data, var, by) {
  data |>
    group_by({{by}}) |>
    summarise(maximum = max({{var}}, na.rm = TRUE))
}
```

-   Infix Operators
    -   Most infix operators (`==`, `+`, `-`, `<-`, etc.) should always be surrounded by spaces.  
    -   There are a few exceptions, which should never be surrounded by spaces:
        -   Operators with high precendence (`::`, `:::`, `$`, `@`, `[`, `[[`, `^`, `:`), single-sided formulas when the right-hand side is a single identified, or when used in tidy evaluation `!!` (bang-bang) and `!!!` (bang-bang-bang).  
        
```{r, eval=FALSE}
# Good
height <- (feet * 12) + inches
mean(x, na.rm = TRUE)

sqrt(x^2 + y^2)
df$z
x <- 1:10

~foo
tribble(
  ~col1, ~col2,
  "a",   "b"
)

call(!!xyz)

# Bad
height<-feet*12+inches
mean(x, na.rm=TRUE)

sqrt(x ^ 2 + y ^ 2)
df $ z
x <- 1 : 10

~ foo
tribble(
  ~ col1, ~ col2,
  "a", "b"
)

call(!! xyz)
call( !! xyz)
call(! !xyz)
```

