---
title: "Using Git and GitHub/GitLab"
format: docx
---

This is a quick guide to using Git and GitHub in R 
following [Happy Git and GitHub for the useR](https://happygitwithr.com/).  

## Setup

- Register a [GitHub](https://github.com) account if you don't already have one.  
- Install [Git](https://git-scm.com/install/)  

## Introduce yourself to Git  

**Using the shell**  

```{r}
#| eval: false

git config --global user.name "Jane Doe"
git config --global user.email "github-account-email@fake.com"
git config --global --list
```

**Using the `usethis` package in R**  

```{r}
#| eval: false
# install.packages("usethis")
usethis::use_git_config(
  use.name = "Jane Doe", 
  user.email = "github-account-email@fake.com"
)
```

### Configuring the default name for the initial branch

You may want to configure the default branch name for the new repo. 
Previously, the default branch was named `master` but you can change this 
using the shell or `usethis` package.

```{r}
#| eval: false

# shell
git config --global init.defaultBranch main
# usethis
usethis::git_default_branch_configure() # default: name = "main"
```

## Connect Git, GitHub, RStudio

Setting up the connection for the HTTPS protocol using a **personal access token (PAT)**.
